---
title: "Data Visualisation with ggplot2 - Project"
output: html_notebook
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the reasearch question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?
## the data is relevant

## 2. Reading the Data

```{r}
# Load the data below
# --- 
library(tidyverse)
# Dataset url = https://bit.ly/2WvQbhJ

# --- 
# YOUR CODE GOES BELOW
# 
hotelbookings_df<- read.csv("hotel_bookings.csv")
hotelbookings_df
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(hotelbookings_df,5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(hotelbookings_df,5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample(hotelbookings_df,10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotelbookings_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(hotelbookings_df)
```

## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
str(hotelbookings_df)
colSums(is.na(hotelbookings_df))
```


```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
hotelbookings_df[duplicated(hotelbookings_df), ]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 
colSums(is.na(hotelbookings_df))
```


```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
# 
is.na(hotelbookings_df)
```

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 
install.packages("reshape2")
library(reshape2)
# Select the desired columns
hotelbookings_df <- hotelbookings_df %>%
na.omit() %>%
  select()
hotelbookings_df
# Create a correlation matrix
corr_matrix <- cor(hotelbookings_df, method="s")
head(corr_matrix)
# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df
# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```
###
###

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
hotelbookings_df2 = select(hotelbookings_df, -c(company))
names(hotelbookings_df2)
```

From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#
hotelbookings_df$distribution_channel[is.na(hotelbookings_df$distribution_channel)] <- "TO"
hotelbookings_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#
adults <-hotelbookings_df2$adults
babies <-hotelbookings_df2$babies
child <-hotelbookings_df2$children
hotelbookings_df3 <- subset(Bookings_df2, adults!="0" & babies!="0" & child!="0")
head(Bookings_df3)

```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
hotelbookings_df3$children[is.na(hotelbookings_df3$children)] <- mean(dt$age, na.rm = TRUE)
hotelbookings_df3
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
is.na(hotelbookings_df3)
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
lapply(hotelbookings_df3$children,as.numeric)
lapply(hotelbookings_df3$company,as.numeric)
lapply(hotelbookings_df3$agent,as.numeric)
glimpse(hotelbookings_df3)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot
```
booking_cancelled_df <- hotelbookings_df3 %>%
  group_by(hotel) %>%
  summarise(is_canceled = sum(is_canceled))
# Plotting Barplot chart to visualize the data
ggplot(booking_cancelled_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Cancelled Bookings", 
    x = "Hotels", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = is_canceled, face = "bold"), vjust = 1.5, colour = "black")

```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel
booking_ratio_df <- hotelbookings_df3 %>%
  group_by(hotel) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(ratio_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "booking ratio between resort hotel and city hotel", 
    x = "Type of Hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# 
year_df <- Bookings_df %>%
  group_by(arrival_date_year) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(year_df, aes(x = arrival_date_year, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Year", 
    x = "year", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#
countries_df <- Bookings_df %>%
  group_by(country) %>%
  summarise(is_canceled = sum(is_canceled==0))
countries_df
arrange(countries_df,desc(is_canceled))
#Filtering to get the top three countries
top3_countries_df <- filter(countries_df, country == "PRT" | country == "GBR" | country == "FRA")
top3_countries_df
# Plotting Barplot chart to visualize the data
ggplot(top3_countries_df, aes(x = country, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "country", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 
stays_df <- Bookings_df %>%
  mutate(total_stays = stays_in_weekend_nights + stays_in_week_nights) %>%
  group_by(total_stays) %>%
  summarise(is_canceled = sum(is_canceled==0))
  arrange(stays_df, desc(is_canceled))
stays_df
# Plotting Barplot chart to visualize the data
ggplot(stays_df, aes(x = total_stays)) +
  geom_histogram(
    bins = 10,fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Total stays in nights", 
    x = "total_stays", 
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  scale_x_continuous(breaks = seq(0, 50,by=5))
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation_df <- Bookings_df %>%
  mutate(hotel_accommodation =
           case_when(
             Bookings_df$adults >= 0 & (Bookings_df$children >= 1 | Bookings_df$babies >= 1) ~ "Family",
             Bookings_df$adults == 2 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Couple",
             Bookings_df$adults == 1 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Single",
            Bookings_df$adults > 2 & (Bookings_df$children == 0 | Bookings_df$babies == 0) ~ "Family")
  )%>%
  filter(is_canceled == 0)
  
accommodation_df1 <- accommodation_df %>%
  group_by(hotel_accommodation) %>%
  count(hotel_accommodation)
#Plotting the bar chart
ggplot(accommodation_df1, aes(x = hotel_accommodation, y = n, fill = accommodation_df)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Hotel Accommodataion by accomodation type",
    x = "Hotel Accommodation", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

Our observations:
-

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
b). Do we need other data to answer our question?
c). Did we have the right question?


